---
title: "Day 7 - Activities using joins, dates, string, and case when"
author: "Colleen McCamy"
date: "2022-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(tidyverse)
library(lubridate)

```


Read in the data sets and store the data frames as bird_observations, sites, surveys, and taxalist (it should be clear from the raw file names which is which)

```{r}

bird_observations <- read_csv(here::here("data", "52_pp52_birds_1.csv"))

sites <- read_csv(here::here("data", "52_pp52_sites_1.csv"))

surveys <- read_csv(here::here("data", "52_pp52_surveys_1.csv"))

taxalist <- read_csv(here::here("data", "52_pp52_taxalist_1.csv"))
```

Create a subset of bird_observations called birds_subset that only contains observations for birds with species id “BHCO” and “RWBL”, and from sites with site ID “LI-W” and “NU-C”

```{r}

birds_subsets <- bird_observations |> 
  filter(species_id %in% c("BHCO", "RWBL")) |> 
  filter(site_id %in% c("LI-W", "NU-C"))

```

## Left Join Practice
Use left join(s) to update birds_subset so that it also includes sites and taxalist information. For each join, include an explicit argument saying which variable you are joining by (even if it will just assume the correct one for you). Store the updated data frame as birds_left. Make sure to look at the output - is what it contains consistent with what you expected it to contain?

```{r}

birds_left <- left_join(birds_subsets, sites, by = "site_id") |> 
  left_join(taxalist, by = "species_id")

# below is another way to write it that makes it more intuitive to me

birds_left_2 <- birds_subsets |> 
  left_join(sites, by = "site_id") |> 
  left_join(taxalist, by = "species_id")
```

## Full Join Practice

First, answer: what do you expect a full_join() between birds_subset and sites to contain? 

Answer - all data will be kept from both the data sets and it will be matched on the sites 

Write code to full_join the birds_subset and sites data into a new object called birds_full. Explicitly include the variable you’re joining by. Look at the output. Is it what you expected?


```{r}

birds_full <- birds_subsets |> 
  full_join(sites, by = "site_id")

# yes, I can see that the columns from sites are included in birds_subset

```

## Inner Join Practice
First, answer: what do you expect an inner_join() between birds_subset and taxalist to contain? 

Answer - only species that are in the taxalist will be in the bird_subset - however sicne all of the species in the subset are in the taxalist I think that it should be the same as bird subset

Write code to inner_join() the birds_subset and taxalist, called birds_inner. Include an argument for what variable you’ll be joining by.

```{r}

birds_inner_join <- inner_join(birds_subsets, taxalist, by = "species_id")
birds_left_join <- left_join(birds_subsets, taxalist, by = "species_id")

```

What would you get if instead of inner_join() you’d used left_join() for this example? Try it out & check. Why does that make sense for this scenario? It what case would you expect the outcome to differ from an inner_join()?

Answer - they should be the same since all birds in the bird subset have an observation in the taxalist species ID code thus, doing left_join or inner_join makes no difference

## Task 2: Data wrangling and visualization with birds

Starting with your birds object, rename the notes column to bird_obs_notes (so this doesn’t conflict with notes in the surveys dataset

```{r}

bird_rename <- bird_observations |> 
  rename("bird_obs_notes" = notes)

```

Then, create a subset that contains all observations in the birds dataset, 
joins: 
-the taxonomic, 
-site and 
-survey information to it, and 

is limited to only columns survey_date, common_name, park_name, and bird_count. 

You can decide the order that you want to create this in (e.g. limit the columns first, then join, or the other way around).

```{r}

birds_new_subset <- bird_rename |> 
  left_join(taxalist, by = "species_id") |> 
  left_join(sites, by = "site_id") |> 
  left_join(surveys, by = c("site_id", "survey_id"))

# since we wanted to match for two column variables we joined this by all of the other items above and not only does the survey id have to match but also the site_id

birds_select <- birds_new_subset |> 
  select("survey_date", "common_name", "park_name", "bird_count")

```

Use lubridate::month() to add a new column called survey_month, containing only the month number. Then, convert the month number to a factor (again within mutate())

```{r}


birds_factor <- birds_select |> 
  mutate(survey_date = as.factor(survey_date)) |> 
  mutate(survey_date = lubridate::ymd(survey_date))

birds_factor <- birds_select |> 
  mutate(survey_month = lubridate::month(survey_date)) |> 
  dplyr::relocate(survey_month, .after = survey_date)

# Learn a new function on your own! Use dplyr::relocate() to move the new survey_month column to immediately after the survey_date column. You can do this in a separate code chunk, or pipe straight into it from your existing code.

```

Find the total number of birds observed by park and month (i.e., you’ll group_by(park_name, survey_month))


```{r}

birds_summary <- birds_factor |> 
  group_by(park_name, survey_month) |> 
  summarize(total_birds_obs = sum(bird_count, na.rm = TRUE))

# Filter to only include parks “Lindo”, “Orme”, “Palomino” and “Sonrisa”

birds_summary_filter <- birds_summary |> 
  filter(park_name %in% c("Lindo", "Orme", "Palomino", "Sonrisa"))

```




